---
author: "Homesh Wathsalya Ampe Mohottige"
date: "15/08/2021"
output: pdf_document
Special Note: Please change the 'data_location' variable, before running the code. Currently the csv data file sits at the same folder as the .rmd file
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
2. Read the dataset into \textbf{R}. Check if there are missing values (NA) and, in case there are, remove them.
```{r q_2, echo=FALSE}
# data file location
data_location <- "winequality-red.csv"

# read the csv to a data frame
wine_df <- read.csv(data_location,stringsAsFactors = F)
nrow(wine_df)

# to check the structure of the dataset
#str(wine_df)

# remove NA values
wine_df <- wine_df[complete.cases(wine_df), ]
nrow(wine_df)

```

\textbf{Comment}
Number of rows before and after removing NAs are similler. Hence no NA values were there in the dataset.\

3. We want to implement a logistic regression, therefore we want a response variable which assume values either 0 or 1. Suppose we consider "good" a wine with quality above 6.5 (included).

```{r q_3, echo=FALSE}
wine_df$quality <- ifelse(wine_df$quality<6.5,0,1)
#head(wine_df)

```
4. Run a frequentist analysis on the logistic model, using the \textbf{glm()} function. What are the significant coefficients?

```{r q_4, echo=FALSE}

obj_log <- glm(wine_df$quality ~ .,family = binomial(link="logit") , data=wine_df)
summary(obj_log)

```

Considering the common $\alpha$ level of 0.05, we can say that following coefficients are significant\

fixed.acidity\
volatile.acidity\
residual.sugar\
chlorides\
total.sulfur.dioxide\
density\
sulphates\
alcohol\

(Note: result implies \textbf{Intercept} is also significant)\

5. Estimate the probabilities of having a "success": fix each covariate at its mean level, and compute the probabilities for a wine to score "good" varying \textbf{total.sulfur.dioxide}, and plot the results.

```{r q_5, echo=FALSE}

# preparing new set of values for total.sulfur.dioxide, to do the predictions
total.sulfur.dioxide.range <- seq(from=min(wine_df$total.sulfur.dioxide), 
                         to=max(wine_df$total.sulfur.dioxide), by=1)

# get mean values for each category (without 'total sulfur dioxide')
meanvals <- colMeans(wine_df[c(1:6,8:11)])
#meanvals
# adding 1 into the vector for the intercept
meanvals <- append(1,meanvals)
meanvals <- as.numeric(meanvals)

# preparing coefficient vector to multiply with mean values (removing "total sulfur dioxide" coefficient)
coeffs <- obj_log$coefficients[c(1:7,9:12)]
coeffs <- as.numeric(coeffs)

# "total sulfur dioxide" coefficient
b_total.sulfur.dioxide <- obj_log$coefficients[c(8)]

# calculating sum of all fixed values
all_fixed_vals_vector <- meanvals*coeffs
sum_of_fixed_vals <- sum(all_fixed_vals_vector)

# calculating response value for each 'total sulfer dioxide' value
quality_val <- sum_of_fixed_vals + b_total.sulfur.dioxide*total.sulfur.dioxide.range

quality_probs <- exp(quality_val)/(1 + exp(quality_val))

plot(total.sulfur.dioxide.range, quality_probs, 
     ylim=c(0,1),
     type="l", 
     lwd=3, 
     lty=2, 
     col="red", 
     xlab="Total Sulfur Dioxide", ylab="P(Quality)", main="Probability of quality prediction")



```

6. Perform a Bayesian analysis of the logistic model for the dataset, i.e. approximate the posterior distributions of the regression coefficients, following these steps: 
6.1 Write an \textbf{R} function for the log posterior distribution.
```{r q_6.1, echo=FALSE}

lpost.LR <- function(beta,x,y)
{
  eta <- as.numeric(x %*% beta)
  logp <- eta - log(1+exp(eta))
  logq <- log(1-exp(logp))
  logl <- sum(logp[y==1]) + sum(logq[y==0])
  lprior <- sum(dnorm(beta,0,10,log=T))
  return(logl + lprior)
}

```

6.2 Fix the number of simulation at 10^4 
```{r q_6.2, echo=FALSE}

# S - Number of simulations
S <- 10^4

```

6.3 Choose 4 different initialisations for the coefficients.
```{r q_6.3, echo=FALSE}

intcpt <- rep(1,nrow(wine_df))
X <- data.matrix(cbind(intcpt,wine_df[c(1:11)]))
Y <- data.matrix(wine_df[c(12)])

# (total columns) - (response var col) + (intercept col)
coeff_count <- ncol(X)

# array to keep the initial value sets for later use
beta_ary <- array(NA, dim=c(4,S,coeff_count))

# values from frequentist linear model
beta_ary[1,1,] <- as.numeric(coefficients(obj_log))

# all 0 coefficients
beta_ary[2,1,] <- as.numeric(rep(0,coeff_count))

# all 0.5 coefficients
beta_ary[3,1,] <- as.numeric(rep(0.1,coeff_count))

# random 11 values from prior distribution
beta_ary[4,1,] <- as.numeric(rep(0.0001,coeff_count))

```
\textbf{Comment}

Initialization 1 - coefficient values from the frequentist analysis\
Initialization 2 - all 0 coefficients\
Initialization 3 - all 0.1 coefficients\
Initialization 4 - all 0.0001 coefficients\


6.4 For each initialisation, run a Metropolis–Hastings algorithm.
```{r q_6.4, echo=FALSE}

# library for multivariate normal function
library(mvtnorm)

Omega_prop <- solve(t(X) %*% X)
k <- coeff_count


for(i in 1:4){
  
  cat("start: algorithm running for initialization - ",i,"\n")
  
  # matrix to remporary store the beta values
  beta_mat <- matrix(NA,nrow=S,ncol=coeff_count)
  beta_mat[1,] <- as.numeric(beta_ary[i,1,])
  
  # variable to calculate acceptance probability
  acc <- 0
  
  for(iter in 2:S)
  {
    flag <- TRUE
    # Propose a new set of values
    beta_star <- rmvnorm(1,beta_mat[iter-1,],0.5*Omega_prop)

    # Compute the posterior density on the proposed value and on the old value  
    newpost<-lpost.LR(t(beta_star),X,Y)
    oldpost<-lpost.LR(matrix(beta_mat[iter-1,],ncol=1),X,Y)
    tryCatch(
      expr={
        # Acceptance step
        if(runif(1,0,1)>exp(newpost-oldpost))
        {
          beta_mat[iter,]=beta_mat[iter-1,]
        } 
        else
        {
          beta_mat[iter,]=beta_star
          acc=acc+1
        }
      },
      # print debug information if the chain outputs infinite values
      error = function(e){
        print("error")
        print(e)
        print(beta_star)
        print(newpost)
        print(oldpost)
        flag<<-FALSE
      }
    )
    # break the loop if chain outputs infinite values
    if(!flag) break
    
  }
  beta_ary[i,,]<-beta_mat
  
  cat("acceptance probability ", acc/iter,"\n")
  
  cat("end: algorithm running for initialization - ",i,"\n")
}


```

6.5 Plot the chains for each coefficients (the 4 chains on the same plot) and comment.

20% burn-in has been chosen\

```{r q_6.5, echo=FALSE}

headings <- append(c("intercept"),names(wine_df[c(1:11)]))

#par(mfrow=c(4,3))
for(j in 1:12){
  #print(obj_log$coefficients[j])
  ylimit <- c(min(beta_ary[,2000:10000,j]),max(beta_ary[,2000:10000,j]))
  plot(beta_ary[1,2000:10000,j],type="l", col="blue",ylab=bquote(beta[.(j-1)]),ylim=ylimit,main=headings[j])
  abline(h=obj_log$coefficients[j],col="red",lty=2)
  par(new=TRUE)
  plot(beta_ary[2,2000:10000,j],type="l", col="green",ylab='',ylim=ylimit)
  par(new=TRUE)
  plot(beta_ary[3,2000:10000,j],type="l", col="orange",ylab='',ylim=ylimit)
  par(new=TRUE)
  plot(beta_ary[4,2000:10000,j],type="l", col="grey",ylab='',ylim=ylimit)
  
}

```
\textbf{Legend}

blue - coefficient values from the frequentist analysis\
green - all 0 coefficients\
orange - all 0.1 coefficients\
gray - all 0.0001 coefficients\

\textbf{Comments}

intercept and density has changed significantly from the frequentist analysis. But these values are converging well.\
'Fixed acidity', 'PH' and 'alcohol'is also not converging to the values obtained form the frequentist analysis (but in lesser magnitudes)\
'free sulpher dioxide' and 'total sulpher dioxide' seems converging to frequentist analysis values\

All initializations seems to have an acceptace probability around 30%, which is higher than 23%, but still acceptable.\




7. Approximate the posterior predictive distribution of an unobserved variable characterised by\

fixed acidity: 7.5\
volatile acidity: 0.6\
citric acid: 0.0\
residual sugar: 1.70\
chlorides: 0.085\
free sulfur dioxide: 5\
total sulfur dioxide: 45\
density: 0.9965\
pH: 3.40\
sulphates: 0.63\
alcohol: 12\

Plot the approximate posterior predictive distribution.\

```{r q_7, echo=FALSE}

# prediction using the first initialization
for(i in 1:4){
  
  y_new <- c(1)
  x_new <- c(1,7.5,0.6,0.0,1.70,0.085,5,45,0.9965,3.4,0.63,12)
  
  beta_mat <- as.matrix(beta_ary[i,,])

  for(iter in 2:S)
  {
    p_new <- exp(sum(beta_mat[iter,] * x_new) ) / (1 + exp(sum(beta_mat[iter,] * x_new) ))
    y_new[iter] <- rbinom(1,1,prob=p_new)
  }
  
  y_new.feq <- table(y_new)
  barplot(y_new.feq, xlab="quality - prediction", ylab="frequency",main=bquote(initialization[.(i)]))

  
}


```
\textbf{Comments}

Initialization from 0.1 values seems to have a lower probability to have 0 compared to other 3\

8. Use the \textbf{metrop()} function available in the \textbf{mcmc} package to perform the same analysis on the posterior distribution you have approximated for Question 6. Choose again 10^4 simulations and compare the results with the results obtained with your code. (Here a visual comparison of the chains is enough to get full mark). 

```{r q_8, echo=FALSE}
library(mcmc)

lpost <- function(beta,x,y)
{
  eta <- as.numeric(x %*% beta)
  logp <- eta - log(1+exp(eta))
  logq <- log(1-exp(logp))
  logl <- sum(logp[y==1]) + sum(logq[y==0])
  lprior <- sum(dnorm(beta,0,10,log=T))
  return(logl + lprior)
}


  out_1 <- metrop(lpost,x=X,y=Y, beta_ary[1,1,],10000,scale=0.5*Omega_prop)
  out_2 <- metrop(lpost,x=X,y=Y, beta_ary[2,1,],10000,scale=0.5*Omega_prop)
  out_3 <- metrop(lpost,x=X,y=Y, beta_ary[3,1,],10000,scale=0.5*Omega_prop)
  out_4 <- metrop(lpost,x=X,y=Y, beta_ary[4,1,],10000,scale=0.5*Omega_prop)


for(i in 1:12){
  ylimit <- c(min(out_1$batch[2000:10000,i],out_2$batch[2000:10000,i],out_3$batch[2000:10000,i],out_4$batch[2000:10000,i])
              ,max(out_1$batch[2000:10000,i],out_2$batch[2000:10000,i],out_3$batch[2000:10000,i],out_4$batch[2000:10000,i]))
  plot(out_1$batch[2000:10000,i],type="l", col="blue",ylab=bquote(beta[.(i-1)]),ylim=ylimit, xlim=c(0,8000),main=headings[i])
  par(new=TRUE)
  plot(out_2$batch[2000:10000,i],type="l", col="green",ylab=bquote(beta[.(i-1)]),ylim=ylimit, xlim=c(0,8000))
  par(new=TRUE)
  plot(out_3$batch[2000:10000,i],type="l", col="orange",ylab=bquote(beta[.(i-1)]),ylim=ylimit, xlim=c(0,8000))
  par(new=TRUE)
  plot(out_4$batch[2000:10000,i],type="l", col="grey",ylab=bquote(beta[.(i-1)]),ylim=ylimit, xlim=c(0,8000))
}

```
\textbf{Legend}

Blue - coefficient values from the frequentist analysis\
Green - all 0 coefficients\
Orange - all 0.1 coefficients\
Grey - all 0.0001 coefficients\

\textbf{Comments}

Again, only intercept and density have a proper convergence to the same value from all initializations.\
Initialization with 0.1 for each coeficient, does not perform well in fixed acidity and PH. Seems not converging\
Most values are converging, but depending on the initializations, there are considerable differences what we havent seen in question 6.\
  For example in 'volatile acid', 'chlorides', 'suphates' and alchohol, chains initialized with frequentist results, seems to be converging into a totally different values compared to the other initializations.\
Initialization of 0.1 has shown not convergence in several coefficients such as 'chlorides', 'cirtric acid', 'ph', 'fixed acidity' etc\

Overall, the use of metrop() function has proven less effective for some initializations, compared to manually executing the chains.\

























